A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-03-04 pkg/colorspace/paper/Toolbox/article.Rmd
added civids palette to divergingx_hcl and vignettes/article
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@461 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

FIXME: More comparison with/discussion of other palette packages.
the viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
as provided by \proglang{R} package \pkg{viridis} [@color:viridis], and
the scientific color maps of @color:Crameri:2018 as provided by \proglang{R}
package \pkg{scico} [@color:scico]. All yield mostly well-balanced color palettes
and can be approximated reasonably well in \pkg{colorspace}
(see Section\ \ref{sec:approximations}).
A broad range of packages that provide color palettes that are "in harmony" as
they are extracted and clustered from movies, paintings, etc. However, many of these
are not so suitable for scientific visualization as they are not clearly 
qualitative/sequential/diverging. An influential package following such an
approach is \pkg{weanderson} [@color:wesanderson]. The \pkg{paletteer}
[@color:paletteer] package aggregates these and provides them in a unified
user interface.
Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
coordinates, along with visualization and assessment of palettes (beyond
color swatches).

@@

## Flexible diverging palettes

@@

* `"ArmyRose"` to `"Tropic"` closely match the palettes of the same name from \pkg{CARTO} [@color:CARTO].
* `"PuOr"` to `"Spectral"` closely match the palettes of the same name from \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003].

@@

In Figure\ \ref{fig:brewer-carto-viridis} `specplot()` panels (see also
Section\ \ref{sec:specplot}) are shown for two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
Each panel compares the hue, chroma, and luminance trajectories
of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).
The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
admissible RGB colors.

@@

```{r brewer-carto-viridis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of four palettes taken from \\pkg{ColorBrewer.org}, \\pkg{CARTO}, and \\pkg{viridis} (top swatch, solid lines) along with their HCL-based approximations (bottom swatch, dashed lines)."}
specplot(
  rev(RColorBrewer::brewer.pal(7, "YlGnBu")),
  sequential_hcl(7, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
specplot(
  viridis::viridis(7),
  sequential_hcl(7, "Viridis"),
  main = "viridis: Viridis"
)
specplot(
  rcartocolor::carto_pal(7, "ag_Sunset"),
  sequential_hcl(7, "ag_Sunset"),
  main = "CARTO: ag_Sunset"
)
specplot(
  viridis::plasma(7),
  sequential_hcl(7, "Plasma"),
  main = "viridis: Plasma"
)
```

>>>>

FIXME: More comparison with/discussion of other palette packages.
the viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
as provided by \proglang{R} package \pkg{viridis} [@color:viridis], and
the scientific color maps of @color:Crameri:2018 as provided by \proglang{R}
package \pkg{scico} [@color:scico]. All yield mostly well-balanced color palettes
and can be approximated reasonably well in \pkg{colorspace}
(see Section\ \ref{sec:approximations}).
A broad range of packages that provide color palettes that are "in harmony" as
they are extracted and clustered from movies, paintings, etc. However, many of these
are not so suitable for scientific visualization as they are not clearly 
qualitative/sequential/diverging. An influential package following such an
approach is \pkg{wesanderson} [@color:wesanderson]. The \pkg{paletteer}
[@color:paletteer] package aggregates these and provides them in a unified
user interface.
Aims of code in \pkg{colorspace} are general trajectories, adaptable HCL
coordinates, along with visualization and assessment of palettes (beyond
color swatches).

@@

## Flexible diverging palettes {#sec:divergingx}

@@

* `"ArmyRose"` to `"Tropic"` closely match the palettes of the same name from \pkg{CARTO} [@color:CARTO].
* `"PuOr"` to `"Spectral"` closely match the palettes of the same name from \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003].
* `"Cividis"` closely matches the palette of the same name from the \pkg{viridis} family [@color:viridis].
  Note that despite having two "arms" with blue vs.\ yellow colors and a low-chroma center color,
  this is probably better classified as a sequential palette due to the monotonic chroma going from dark to light.
  (See Section\ \ref{sec:approximations} for more details.)

@@

```{r brewer-carto-viridis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of four palettes taken from \\pkg{ColorBrewer.org}, \\pkg{CARTO}, and \\pkg{viridis} (top swatches, solid lines) along with their HCL-based approximations (bottom swatches, dashed lines)."}
specplot(
  rev(RColorBrewer::brewer.pal(7, "YlGnBu")),
  sequential_hcl(7, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
specplot(
  viridis::viridis(7),
  sequential_hcl(7, "Viridis"),
  main = "viridis: Viridis"
)
specplot(
  rcartocolor::carto_pal(7, "ag_Sunset"),
  sequential_hcl(7, "ag_Sunset"),
  main = "CARTO: ag_Sunset"
)
specplot(
  viridis::plasma(7),
  sequential_hcl(7, "Plasma"),
  main = "viridis: Plasma"
)
```

Figure\ \ref{fig:brewer-carto-viridis} shows a selection of such approximations using `specplot()` 
(see also Section\ \ref{sec:specplot}) for two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
Each panel compares the hue, chroma, and luminance trajectories
of the original palettes (top swatches, solid lines) and their HCL-based approximations (bottom swatches, dashed lines).
The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
admissible RGB colors.

@@

```{r cividis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of \\texttt{viridis::cividis} (top swatch, solid lines) along with an HCL-based approximation (bottom swatch, dashed lines)."}
specplot(viridis::cividis(9), divergingx_hcl(9, "Cividis"), main = "viridis: Cividis")
```

Another somewhat nonstandard palette from the viridis family is the
cividis palette based on blue and yellow hues and hence safe for red-green deficient viewers.
Figure\ \ref{fig:cividis} shows the corresponding `specplot()` along with an HCL-based approximation.
What is unusual about this palette: The hue and chroma trajectories would suggest a diverging palette,
as there are two "arms" wth different hues and a zero-chroma point in the center. However,
the luminance trajectory clearly indicates a sequential palette as colors go monotonically from
dark to light. Due to this unusual mixture the palette cannot be composed using the trajectories from
Table\ \ref{tab:hcl}.

However, the tools in \pkg{colorspace} can still be employed to easily reconstruct the palette.
One strategy would be to set up the trajectories manually, using a linear luminance, piecewise
linear chroma, and piecewise constant hue:

```{r cividis_hcl}
cividis_hcl <- function(n) {
  i <- seq(1, 0, length.out = n)
  hex(polarLUV(
    L = 92 - (92 - 13) * i,
    C = approx(c(1, 0.9, 0.5, 0), c(30, 50, 0, 95), xout = i)$y,
    H = c(255, 75)[1 + (i < 0.5)]
  ), fix = TRUE)
}
```

Instead of constructing the hex code from the HCL coordinates via \pkg{colorspace}'s `hex(polarLUV(L, C, H))`,
the base \proglang{R} function `hcl(H, C, L)` from \pkg{grDevices} could also be used.

In addition to manually setting up a dedicated function `cividis_hcl()`, it is possible to approximate the
palette using `divergingx_hcl()` (see Section\ \ref{sec:divergingx}), e.g.,

```{r diverging_cividis, eval = FALSE}
divergingx_hcl(n,
  h1 = 255, h2 = NA, h3 = 75,
  c1 = 30, cmax1 = 47, c2 = 0, c3 = 95,
  l1 = 13, l2 = 52, l3 = 92,
  p1 = 1.1, p3 = 1.0
)
```

This uses a slight power transformation with `p1 = 1` in the blue arm of the palette but otherwise essentially
corresponds to what `cividis_hcl()` does. For convenience `divergingx_hcl(n, palette = "Cividis")` is preregistered
using the above parameters.
